### 2021-07-01 rad-icfp-2021/background.tex
minor corrections intro

The two main ways in which it is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
Our work focus on a compiled version so we focus on source code transformation which is more fitted in this case.

@@

Because the compute flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
%%TODO: from Matthijs' paper, need to rewrite.
Reverse-mode AD is only well-understood as a source-code transformation (also called define-then-run
style AD) on limited programming languages. Typically, its implementations
on more expressive languages that have features such as higher-order functions and conditionals
make use of define-by-run approaches. These approaches first build a computation graph during runtime, effectively evaluating the program until a straight-line
first-order program is left, and then they evaluate this new program \cite{carpenter2015stan,paszke2017automatic}. 
Such approaches have the severe downside that the differentiated code cannot benefit from existing optimizing compiler architectures. As such, these AD libraries
need to be implemented using carefully, manually optimized code, that for example does not contain any common subexpressions. This implementation process
is precarious and labour intensive. Further, some whole-program optimizations
that a compiler would detect go entirely unused in such systems.

@@

Let the term we want to differentiate be $x1:\RR,...,xn:\RR\vdash exp(cos(xi))$.
Following the chain rule, we need the jacobian matrices of $cos$ at $xi$ and of $exp$ at $cos(xi)$. 
Instead of considering these as operations from $\RR\to\RR$, we consider then as functions from the whole context. So $cos$ and $exp$ are seen as functions $\RR^n\to\RR$.
If we just do that though, we lose compositionality. 
So $cos$ will also return its context and is now seen as a function $\sem{cos}:\RR^n\to\RR^{n+1}$.
similarly for $exp$, but also taking the return value of $\sem{cos}$ as an extra argument, the one it will actually use and not simply return. 
We thus obtain $\sem{exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $J\sem{cos} \in Mat_{n,n+1}$, $J\sem{exp} \in Mat_{n+1,n+2}$ also compose nicely.
The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ...\circ f_1$. 
If the original program was of type $\RR$, then the return value of the original program is the last compononent of $f$.
Following the mathematical presentation of reverse-mode above, the gradient of the the original program is then obtained as 
\[\nabla f= J^Tf(0,...,0,1)=J^Tf_1(J^Tf_{2}(...(J^Tf_m(0,...,0,1)))...)\]

@@

This method is quite inefficient though. First, we have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
Second, each $J^Tf_{i+1}$ is a potentially huge matrix if $n$ or $m$ is big.

The imperative version of reverse mode for a binary operator $op(x,y)$ is something like $x'+= \partial_1op(x,y);y'+= \partial_2op(x,y)$. 
It is for these cases that mutation is usually key. 
If we see our term as a directed graph, reverse mode needs to backpropagate from the end of the graph to the starting nodes via every path.
It's hard to keep track of all these information in parallel in a functional way.
One of the key simple ideas that we used was to transform every operator into a unary one. 
This essentially trivialises the compute flow to a line. 
Even if the starting program was a straight line program, 
having non unary operators was a source of inefficiency which justified mutation in the first place.
By returning every variable every time, the problem of using a variable several time does not need to be dealt with via mutation. 
This simple idea of transforming a program into essentially a straight line is what our new intermediate representation unary-form does.

@@

If we look at $J^Tf_{i+1}$, we notice that this function is almost the identity, except at the last row. 
Even on the last row, if the original term was a unary or binary operator like $cos, exp, +, *$, 
the row is zero except for at most 2 indices (1 for unary operators).
This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,...,y(n+i)).(y1,...,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
where $[k]$ means that the element is added at the $k$-th index of the tuple.

We know in advance that such jacobian functions are going to be applied one to another, so we can use partial evaluation to $\beta$-reduce all of these $\lambda$s.
Because each function is almost the identity, we obtain a lot of substitutions of the form $[x/y]$ where both $x$ and $y$ are variables. 
This allows us to drastically reduce the size of $J^Tf$. In fact, for simple programs this is basically enough to get an efficient purely functional reverse derivative transformation.
We develop this idea further for a richer language.

>>>>

The main two ways in which it is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
Our approach focuses on source code transformation which is better fitted for compilation and optimizations.

@@

Because the computation flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
(TODO: paragraph below from Matthijs' paper, need to rewrite.)
Reverse-mode AD is only well-understood as a source-code transformation 
(also called define-then-run style AD) on limited programming languages. 
Typically, its implementations on more expressive languages that have features such as higher-order functions and conditionals
make use of define-by-run approaches. 
These approaches first build a computation graph during runtime, effectively evaluating the program until a straight-line
first-order program is left, and then they evaluate this new program \cite{carpenter2015stan,paszke2017automatic}. 
Such approaches have the severe downside that the differentiated code cannot benefit from existing optimizing compiler architectures. 
As such, these AD libraries need to be implemented using carefully, manually optimized code, that for example does not contain any common subexpressions. 
This implementation process is precarious and labour intensive. 
Further, some whole-program optimizations that a compiler would detect go entirely unused in such systems.

@@

Let the term we want to differentiate be $x1:\RR,...,xn:\RR\vdash exp(cos(xi))$.
Following the chain rule, we need the jacobian matrices of $cos$ at $xi$ and of $exp$ at $cos(xi)$. 
Instead of considering these as operations from $\RR\to\RR$, we consider them as functions from the whole context. So $cos$ and $exp$ are seen as functions $\RR^n\to\RR$.
However, by simply doing this we lose compositionality. 
So $cos$ will also return its context and is now seen as a function $\sem{cos}:\RR^n\to\RR^{n+1}$.
Similarly for $exp$, but also taking the return value of $\sem{cos}$ as an extra argument, the one it will actually use and not simply return. 
We thus obtain $\sem{exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $J\sem{cos} \in Mat_{n,n+1}$, $J\sem{exp} \in Mat_{n+1,n+2}$ also compose nicely.
The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ...\circ f_1$. 
If the original program was of type $\RR$, then the return value of the original program is the last component of $f$.
Following the mathematical presentation of reverse-mode above, the gradient of the the original program is then obtained as 
\[\nabla f= J^Tf(0,...,0,1)=J^Tf_1(J^Tf_{2}(...(J^Tf_m(0,...,0,1)))...)\]

@@

This method is quite inefficient still. First, we have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
Second, each $J^Tf_{i+1}$ is a potentially huge matrix if $n$ or $m$ is big.

The imperative version of reverse mode for a binary operator $op(x,y)$ is something like $x'+= \partial_1op(x,y);y'+= \partial_2op(x,y)$. 
It is for these cases that mutation is usually key. 
If we see our term as a directed graph, reverse mode needs to backpropagate from the end of the graph to the starting nodes via every path.
It's hard to keep track of all these information in parallel in a functional way.
One of the key simple ideas that we used was to transform every operator into a unary one. 
This essentially trivialises the computation flow to a line. 
Even if the starting program was a straight line program, 
having non-unary operators was a source of inefficiency which justified mutation in the first place.
By returning every variable every time, the problem of using a variable several times does not need to be dealt with via mutation. 
This simple idea of transforming a program into essentially a straight line is what our new intermediate representation unary-form does.

@@

If we look at $J^Tf_{i+1}$, we notice that this function is almost the identity, except at the last row. 
Even on the last row, if the original term was a unary or binary operator like $cos, exp, +, *$, 
the row is zero except for at most two indices (one for unary operators).
This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,...,y(n+i)).(y1,...,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
where $[k]$ means that the element is added at the $k$-th index of the tuple.

We know in advance that such jacobian functions are going to be applied one to another and we can use partial evaluation to $\beta$-reduce all of these $\lambda$s.
Because each function is almost the identity, we obtain a lot of substitutions of the form $[x/y]$ where both $x$ and $y$ are variables. 
This allows us to drastically reduce the size of $J^Tf$. In fact, for simple programs, this is basically enough to get an efficient purely functional reverse derivative transformation.
We develop this idea further for a richer language.

### 2021-07-01 rad-icfp-2021/figures/compilation_scheme.tex
minor corrections intro

\[
\begin{tikzcd}
    \Source \ar[r] & \UNFSymbol_1 \ar[r,"\Dsynrevsymbol"] & \UNFSymbol_2 \ar[r] & \Target \ar[r,"\partial ev"] & \Target
\end{tikzcd}
\]

>>>>

\[
\begin{tikzcd}
    \Source \ar[rrr,"\text{efficient }\Dsynrevsymbol (Fig.~\ref{fig:direct_diff_macro})"] \ar[dr,"(Fig.~\ref{fig:source_to_unf})"'] &&& \Target \arrow[r,loop right,"\text{optim }(Fig.~\ref{fig:optim})"] \\
    & Source\UNFSymbol \ar[r,"\Dsynrevsymbol (Fig.~\ref{fig:diff_macro})"] & Target\UNFSymbol \ar[ur,"(Fig.~\ref{fig:unf_to_target})"'] & 
\end{tikzcd}
\]

### 2021-07-01 rad-icfp-2021/intro.tex
minor corrections intro

After some general partial evaluation techniques that will be detailed further in the paper, we obtain     
        \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1 = 0 in!\\
                !let y2 = 0 in!\\
                !let y3 = 0 in!\\
                !let y4 = 0 in!\\
                !let y5 = 1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
        \end{center}  
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
Of course, this term can be further optimised via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}

@@

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where !shift1 [v1,...,vn] = [1,v1,...,v(n-1)]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
!scanl! and !scanr! are respectively scan left and scan right. These are like fold but store all intermediate results in an array.
We obtain these gradients, which are efficient and usually derived by hand, as special cases of our work.
They are obtained automatically in our case.
\end{example}

@@

Our work looks like the following compilation scheme
\input{figures/compilation_scheme}

In Section~\ref{sec:background} we recall rudiments of automatic differentiation, and in particular the idea of reverse-mode and the cheap gradient principle.
Next, in Section~\ref{sec:simplediff} we introduce the source language we want to differentiate. We also introduce our new intermediate representation UNF, and a simple reverse-mode macro on this representation.
In Section~\ref{sec:efficientrad} we translate back from UNF to a purely functional target language and show several general optimisations to make our reverse-mode transformation efficient.
In the next Section~\ref{sec:complexity} we show that the previous optimisations are sufficient to ensure our transformation has the right complexity.
Section~\ref{sec:correctness} shows that our construction is correct and we give standard semantics to our source and target languages. 
In Section~\ref{sec:generalisation} we show how to generalise the previous sections for a richer source language. 
Finally, related work and some limitations to our approach are presentend in Section~\ref{sec:conclusion}.

>>>>

After some general partial evaluation techniques that will be detailed further in the paper, we obtain     
        % \begin{center}
        %     \begin{tabular}{l}
        %         !let w1 = x1 * x2 in!\\ 
        %         !let w2 = w1 * x1 in!\\
        %         !let y1 = 0 in!\\
        %         !let y2 = 0 in!\\
        %         !let y3 = 0 in!\\
        %         !let y4 = 0 in!\\
        %         !let y5 = 1 in!\\
        %         !let y1'! != y1+w1*y5 in!\\
        %         !let y4'! != y4+x1*y5 in!\\
        %         !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
        %     \end{tabular}
        % \end{center}
    \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1,y2,y3,y4,y5 = 0,0,0,0,1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
    \end{center}   
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
This term can be further optimised via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}

@@

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where !shift1 [v1,...,vn] = [1,v1,...,v(n-1)]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
!scanl! and !scanr! are respectively scan left and scan right. These return all the intermediate result of a fold in an array.
These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
\end{example}

@@

Our main transformation is present on the top of the diagram and presented in Section~\ref{sec:simplediff}. 
It it decomposed 3 steps which are presented in Section~\ref{sec:unf}.
\input{figures/compilation_scheme}

In Section~\ref{sec:background} we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
Next, in Section~\ref{sec:simplediff} we introduce the source and target languages for differentiation, and our reverse mode transformation. 
In Section~\ref{sec:unf} we introduce a new intermediate representation UNF, and a simple reverse-mode macro on this representation.
We introduce transformations from the source language to UNF and from UNF to the target language.
Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimisations.
In Section~\ref{sec:generalisation} we show how to adapt our work for a richer source language. 
Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.

### 2021-07-01 rad-icfp-2021/relwork.tex
minor corrections intro

We have shown how to compute provably efficient gradients via reverse derivatives in a purely functional way 
on a simple yet expressive language with array functions.
We introduced a novel intermediate representation, unary form, 
and used it combined with standard functional optimisations to perform our reverse-mode transformation.

>>>>

We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), combined with standard functional optimisations (\ref{fig:optim}).  
We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
to decompose our transformation into simpler and more insightful transformations.
We showed (\ref{sec:generalisation}) how to lift some of the restrictions that
we introduced on arrays and how to extend our approach.
